\documentclass{article}
\usepackage{zmathlang}

\begin{document}

\begin{zed}
[NAME] 
\end{zed}

\begin{zed}
[SURNAME]
\end{zed}

\begin{schema}{TelephoneDirectory}
persons: NAME \fun SURNAME \\ 
phoneNumbers: NAME \pfun TELEPHONE
\where 
\dom phoneNumbers=\dom persons
\end{schema}

\begin{schema}{InitTelephoneDirectory} 
TelephoneDirectory' 
\where 
phoneNumbers' = \{\} \\
persons' = \{\}
\end{schema}

\begin{schema}{AddPerson}
TheTelephoneDirectory \\
name?: NAME \\
surname?: SURNAME \\
\where
name? \mapsto surname? \notin persons\\
persons' = persons \cup \{name? \mapsto surname? \}
\end{schema}

\begin{schema}{AddNumber}
\Delta TelephoneDirectory \\
n?: NAME \\
s?: SURNAME \\
p?: TELEPHONE
\where
n \mapsto s \in persons\\
p? \notin \ran phoneNumbers \\
phoneNumbers' = phoneNumbers \cup \{n \mapsto phone?\}
\end{schema}

\begin{schema}{RemovePerson}
\Delta TelephoneDirectory \\
n?: NAME \\
s?: SURNAME \\
p?: TELEPHONE
\where
n? \mapsto s? \in persons\\
n? \mapsto p? \notin phoneNumbers \\
persons' = persons \setminus \{n? \mapsto s?\}
\end{schema}

\begin{zed}
OUTPUT ::= success | alreadyInDirectory | nameNotInDirectory | numberInUse
\end{zed}

\begin{schema}{Success}
message!:OUTPUT
\where
message! = success
\end{schema}

\begin{schema}{AlreadyInDirectory}
message!:OUTPUT \\
n?: NAME \\
s?: SURNAME
\where
n? \mapsto s \in person \\
message! = alreadyInDirectory
\end{schema}

\begin{schema}{NameNotInDirectory}
message!:OUTPUT \\
n?: NAME \\
s?: SURNAME
\where
n? \mapsto s? \notin persons \\
message! = NotInDirectory
\end{schema}

\begin{schema}{NumberInUse}
message!:OUTPUT \\
p?: TELEPHONE
\where
p? \in \ran phoneNumbers \\
message! = numberInUse
\end{schema}

\begin{zed}
TotalAddPerson \defs (AddPerson \land Success) 
\lor AlreadyInDirectory \\
TotalRemovePerson \defs (RemovePerson \land Success)
 \lor NameNotInDirectory \\
TotalAddNumber \defs (AddNumber \land Success) 
\lor NameNotInDirectory \lor NumberInUse \\
\end{zed}

\end{document}