
\documentclass{article}
\usepackage{zmathlang}
\begin{document}

\begin{zed}
[PERSON]
\end{zed}

\begin{zed}
GENDER::= male | female
\end{zed}

\begin{schema}{GenDB}
parent: PERSON \rel PERSON \\
sex: PERSON \pfun GENDER \\
\where
\dom parent \cup \ran parent \subseteq \dom sex \\
\forall p: PERSON @ p \mapsto p \notin parent\plus \\
\forall p, q, r: PERSON @ \{p \mapsto q, p \mapsto r \} \subseteq parent \\
\land q \neq r \implies sex~q \neq sex~r 
\end{schema}

\begin{schema}{InitGenDB}
  GenDB'
\where
  sex' = \emptyset \\
  parent' =  \emptyset
\end{schema}

\begin{schema}{AddPerson}
\Delta GenDB \\
name?:PERSON \\
morf?: GENDER
\where
name? \notin \dom sex \\
sex' = sex \cup \{name? \mapsto morf?\} \\
parent' = parent
\end{schema}

\begin{schema}{AddRel}
  \Delta GenDB \\
  off?, par? : PERSON
\where
  \{off?, par?\} \subseteq \dom sex \\
  off? \mapsto par? \notin parent \\
  par? \mapsto off? \notin parent \\
  \#(\{off?\} \dres parent) \leq 1 \\
  \forall x:PERSON @ off? \mapsto x \in parent \implies sex~x \neq sex~par? \\
  sex' = sex
\end{schema}

\begin{schema}{ChangeName}
\Delta GenDB \\
old?, new?: PERSON
\where
old? \in \dom sex \\
new? \notin \dom sex \\
sex' = (\{old?\} \ndres sex) \cup \{new? \mapsto sex~old?\} \\
parent' = (\{old?\} \ndres parent \nrres \{old?\}) \\
\cup \{x:PERSON | x \in parent \limg\{old?\} \rimg @ new? \mapsto x \} \\
\cup \{x:PERSON | x \notin parent\inv \limg \{old?\} \rimg @ x \mapsto new?\}
\end{schema}

\begin{schema}{ChangeSex}
\Delta GenDB \\
p?: PERSON 
\where
p? \in \dom sex \\
sex' = sex \oplus \\
(\{ q: PERSON; s:GENDER | \\
((q \in (parent\inv \semi parent)\plus \limg \{p?\} \rimg)
\land (s \neq sex~ q) @ q \mapsto s \}) \\
parent' = parent
\end{schema}

\begin{schema}{CommonAncestors}
\Xi GenDB \\
p?, q?: PERSON \\
cas!: \power PERSON
\where
\{p?, q?\} \cup cas! \subseteq \dom sex \\

cas! = \{ca: PERSON | \exists m, n:\nat @ \\
((p? \mapsto ca \in parent^{n} \land q? \mapsto ca \in parent^{m}) \\
\land \lnot \exists r:PERSON; x, y: \nat @ ((x+y < m + n ) \\
 \land p? \mapsto r \in parent^{x} \land q? \mapsto r \in parent^{y}))\}
\end{schema}

\begin{schema}{Cousins}
\Xi GenDB \\
p?: PERSON \\
nth?: \nat_{1} \\
rem?: \nat \\
cousins!: \power PERSON
\where
\{p?\} \cup cousins! \dom sex \\

\LET cosrel == (parent^{nth?+1} \semi (parent\inv)^{nth?+1+rem?}) \setminus
 (parent \semi parent\inv) @ \\
 cousins! = cosrel \limg\{p?\}\rimg \cup cosrel\inv\limg\{p?\}\rimg
 \end{schema}





\end{document}
