
\documentclass{article}
\usepackage{zmathlang}
\begin{document}

\dratheory{T7}{0.3}{
\begin{zed}
[PERSON]
\end{zed}

\begin{zed}
GENDER::= male | female
\end{zed}

\draschema{SS1}{
\begin{schema}{GenDB}
parent: PERSON \rel PERSON \\
sex: PERSON \pfun GENDER \\
\where
\draline{SI1}{\dom parent \cup \ran parent \subseteq \dom sex \\
\forall p: PERSON @ p \mapsto p \notin parent\plus \\
\forall p, q, r: PERSON @ \{p \mapsto q, p \mapsto r \} \subseteq parent \\
\land q \neq r \implies sex~q \neq sex~r }
\end{schema}}

\requires{SS1}{SI1}

\draschema{IS1}{
\begin{schema}{InitGenDB}
  GenDB'
\where
  \draline{PO1}{sex' = \emptyset \\
  parent' =  \emptyset}
\end{schema}}

\initialof{IS1}{SS1}
\requires{IS1}{PO1}

\draschema{CS1}{
\begin{schema}{AddPerson}
\Delta GenDB \\
name?:PERSON \\
morf?: GENDER
\where
\draline{PRE1}{name? \notin \dom sex} \\
\draline{PO2}{sex' = sex \cup \{name? \mapsto morf?\} \\
parent' = parent}
\end{schema}}

\uses{CS1}{IS1}
\requires{CS1}{PRE1}
\allows{PRE1}{PO2}

\draschema{CS2}{
\begin{schema}{AddRel}
  \Delta GenDB \\
  off?, par? : PERSON
\where
\draline{PRE2}{  \{off?, par?\} \subseteq \dom sex \\
  off? \mapsto par? \notin parent \\
  par? \mapsto off? \notin parent \\
  \#(\{off?\} \dres parent) \leq 1} \\
\draline{PO3}{\forall x:PERSON @ off? \mapsto x \in parent \implies sex~x \neq sex~par? \\
  sex' = sex}
\end{schema}}

\uses{CS2}{IS1}
\requires{CS2}{PRE2}
\allows{PRE2}{PO3}

\draschema{CS3}{
\begin{schema}{ChangeName}
\Delta GenDB \\
old?, new?: PERSON
\where
\draline{PRE3}{old? \in \dom sex \\
new? \notin \dom sex} \\
\draline{PO4}{sex' = (\{old?\} \ndres sex) \cup \{new? \mapsto sex~old?\} \\
parent' = (\{old?\} \ndres parent \nrres \{old?\}) \\
\cup \{x:PERSON | x \in parent \limg\{old?\} \rimg @ new? \mapsto x \} \\
\cup \{x:PERSON | x \notin parent\inv \limg \{old?\} \rimg @ x \mapsto new?\}}
\end{schema}}

\uses{CS3}{IS1}
\requires{CS3}{PRE3}
\allows{PRE3}{PO4}

\draschema{CS4}{
\begin{schema}{ChangeSex}
\Delta GenDB \\
p?: PERSON 
\where
\draline{PRE4}{p? \in \dom sex} \\
\draline{PO5}{sex' = sex \oplus \\
(\{ q: PERSON; s:GENDER | \\
((q \in (parent\inv \semi parent)\plus \limg \{p?\} \rimg)
\land (s \neq sex~ q) @ q \mapsto s \}) \\
parent' = parent}
\end{schema}}

\uses{CS4}{IS1}
\requires{CS4}{PRE4}
\allows{PRE4}{PO5}

\draschema{OS1}{
\begin{schema}{CommonAncestors}
\Xi GenDB \\
p?, q?: PERSON \\
cas!: \power PERSON
\where
\draline{PRE5}{\{p?, q?\} \cup cas! \subseteq \dom sex} \\
\draline{O1}{
cas! = \{ca: PERSON | \exists m, n:\nat @ \\
((p? \mapsto ca \in parent^{n} \land q? \mapsto ca \in parent^{m}) \\
\land \lnot \exists r:PERSON; x, y: \nat @ ((x+y < m + n ) \\
 \land p? \mapsto r \in parent^{x} \land q? \mapsto r \in parent^{y}))\}}
\end{schema}}

\uses{OS1}{SS1}
\requires{OS1}{PRE5}
\allows{PRE5}{O1}

\draschema{OS2}{
\begin{schema}{Cousins}
\Xi GenDB \\
p?: PERSON \\
nth?: \nat_{1} \\
rem?: \nat \\
cousins!: \power PERSON
\where
\draline{PRE6}{\{p?\} \cup cousins! \subseteq \dom sex} \\
\draline{O2}{\LET cosrel == (parent^{nth?+1} \semi (parent\inv)^{nth?+1+rem?}) \setminus \\
 (parent \semi parent\inv) @ \\
 cousins! = cosrel \limg\{p?\}\rimg \cup cosrel\inv\limg\{p?\}\rimg}
 \end{schema}}
 
\uses{OS2}{SS1}
\requires{OS2}{PRE6}
\allows{PRE6}{O2}
}
\end{document}
